---
title: "datacleaning"
output: html_document
date: "2024-12-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data cleaning and collating 
```{r}
##Install tidyverse package
library(tidyverse)

#Define the correct field order from the IMPC_SOP.csv file
sop_file <- "~/Desktop/DCDM_Group7/originals/IMPC_SOP.csv" #path to IMPC_SOP
sop_data <- read.csv(sop_file, header = TRUE) 

# Extract and clean the correct field names (standardized to lowercase)
correct_fields <- tolower(trimws(sop_data[[1]]))  # Lowercase and remove extra spaces

# Set the path to the folder containing all 140k files
data_path <- "~/Desktop/DCDM_Group7/originals/7/"

# List all files in the folder
file_list <- list.files(path = data_path, pattern = "\\.csv$", full.names = TRUE)

# Creata an empty list to store data
combined_data <- list()

```


#Create a loop to process each file 

```{r}
for (file in file_list) {
  # Read the file
  data <- read.csv(file, header = FALSE, col.names = c("field", "value"))
  
  # Clean up the 'field' names: lowercase and remove spaces
  data$field <- tolower(trimws(data$field))
  
  # Transpose the 'value' column so that fields become column names
  reshaped_data <- t(data$value) %>%
    as.data.frame() %>%
    setNames(data$field)
  
  # Ensure all columns are in the correct order
  # Add missing columns with NA
  missing_fields <- setdiff(correct_fields, names(reshaped_data))  # Find fields that are missing
  
  
  # Reorder columns to match the correct field order
  final_data <- reshaped_data %>%
    select(all_of(correct_fields))
  
  # Append to the list
  combined_data[[length(combined_data) + 1]] <- final_data
}
```

# Combine all data frames into one
```{r}
final_combined_data <- bind_rows(combined_data)

# Write the combined data to a single CSV file
write.csv(final_combined_data, "combined_output.csv", row.names = FALSE)
```


# Validate string length ( 15 characters) in analysis id
```{r}
# Define the function 

validate_analysis_id <- function (x, dataType, min_val, max_val) {
  if (tolower(dataType) == "string") {
    invalid_analysis_id <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)
    return(invalid_analysis_id)
  } 
}
  
# Call the function 
  
invalid_analysis_id <- validate_analysis_id(final_combined_data$analysis_id, "string", 15, 15)

#Print results 
print(invalid_analysis_id)

# The returned results, is named_integer (0). This means that all strings are valid, and have the lengths defined in the SOP file. 
```

# Validate string length for gene accession id 
```{r}
# Define the function 
validate_gene_accession_id <- function (x, dataType, min_val, max_val) {
  if (tolower(dataType) == "string") {
    invalid_gene_accession_id <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)
    return(invalid_gene_accession_id)
  } 
}
  
# Call the function 
invalid_gene_accession_id <- validate_gene_accession_id(final_combined_data$gene_accession_id, "string", 9, 11)

#Print results 
print(invalid_gene_accession_id)

# The returned results, is named_integer (0). This means that all strings are valid, and have the lengths defined in the SOP file.

```

# Validate string length for gene_symbol 
```{r}
# Define the function 
validate_gene_symbol <- function (x, dataType, min_val, max_val) {
  if (tolower(dataType) == "string") {
    invalid_gene_symbol <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)
    return(invalid_gene_symbol)
  } 
}
  
# Call the function 
invalid_gene_symbol <- validate_gene_symbol(final_combined_data$gene_symbol, "string", 1, 13)

#Print results 
print(invalid_gene_symbol)

# The returned results, is named_integer (0). This means that all strings are valid, and have the lengths defined in the SOP file.
```

# Validate string length for parameter_id 
```{r}
# Define the function 
validate_parameter_id <- function (x, dataType, min_val, max_val) {
  if (tolower(dataType) == "string") {
    invalid_parameter_id <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)
    return(invalid_parameter_id)
  } 
}
  
# Call the function 
invalid_parameter_id <- validate_parameter_id (final_combined_data$parameter_id, "string", 15, 18)

#Print results 
print(invalid_parameter_id)

# The returned results, is named_integer (0). This means that all strings are valid, and have the lengths defined in the SOP file.
```

# Validate parameter_name 
```{r}
# Define the function 
validate_parameter_name <- function (x, dataType, min_val, max_val) {
  if (tolower(dataType) == "string") {
    invalid_parameter_name <- which(sapply(x, nchar) < min_val | sapply (x, nchar) > max_val)
    return(invalid_parameter_name)
  } 
}
  
# Call the function 
invalid_parameter_name <- validate_parameter_name(final_combined_data$parameter_name, "string", 2, 74)

#Print results 
print(invalid_parameter_name)

# The returned results, is named_integer (0). This means that all strings are valid, and have the lengths defined in the SOP file.
```


# Validate p-value (float) 
```{r}
# Define the function 
validate_p_value <- function (x, min_val, max_val) {
  x <- as.numeric(x)
  invalid_p_value <- which(as.numeric(x) < min_val |
                             as.numeric(x) > max_val)
  return(invalid_p_value)
}
  
# Call the function 
invalid_p_value <- validate_p_value(final_combined_data$pvalue, 0, 1)

#Print results 
#print(invalid_p_value) #this will print the number of rows 

#Print the total number of invalud p-values 
cat("Total number of invalid p-values:", length(invalid_p_value))

# The total entries of invalid p-values is 813. This means that 813 rows are greater than or lower than 0 or 1.  
```

# Validate mouse strain 

```{r}
# Define the function 
validate_mouse_strain <- function (x, field_name) {
  if (field_name == "mouse_strain") {
    valid_mouse_strains <- c("C57BL", "B6J", "C3H", "129SV")
    invalid_mouse_strain <- which(!(x %in% valid_mouse_strains))
    return(invalid_mouse_strain)
  }
}

# Call the function 
invalid_mouse_strain <- validate_mouse_strain(final_combined_data$mouse_strain, "mouse_strain")

#Print the invalid rows 
print(invalid_mouse_strain)

# Print the total number of invalid mouse strains
cat("Total number of invalid mouse strain:", length(invalid_mouse_strain))

# The results show there are 1146 invalid mouse strain entries. There are 1146 strains that do not match the required stages in the SOP file. 

```

# Validate the mouse life stage 

```{r}
# Define the function 
validate_mouse_life_stage <- function (x, field_name) {
  if (field_name == "mouse_life_stage") {
    valid_mouse_life_stage <- c("E12.5", "E15.5", "E18.5", "E9.5", 
      "Early adult", "Late adult", "Middle aged adult")
    invalid_mouse_life_stage <- which(!(x %in% valid_mouse_life_stage))
    return(invalid_mouse_life_stage)
  }
}

# Call the function 
invalid_mouse_life_stage <- validate_mouse_life_stage(final_combined_data$mouse_life_stage, "mouse_life_stage")

#Print the invalid rows 
print(invalid_mouse_life_stage)

# Print the total number of invalid mouse strains
cat("Total number of invalid mouse life stages:", length(invalid_mouse_life_stage))

# The results show there are zero entries with invalid life stages. 
```

#Looking for Duplicates

```{r}
combined_output <- read.csv("~/Desktop/DCDM_GROUP7/combined_output.csv")
combined_output

# Find duplicate rows
duplicates <- duplicated(combined_output)

# View duplicate rows
duplicate_rows <- combined_output[duplicates, ]
print(duplicate_rows)

```

```{r}
# Replace 'column_name' with the name of the column (analysis id)
duplicates_in_column <- combined_output[duplicated(combined_output$analysis_id), ]
print(duplicates_in_column)
```











